---
title: "200206688-MAS6024"
output:
  
  
  bookdown::pdf_document2:
    toc: false
    keep_text: yes
  pdf_document: default
---

```{r setup, include=FALSE, out.height="25%"}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE
                      )
```

Forming the letter grid of the board
```{r, echo = TRUE}
lgrid <- matrix(NA, nrow = 8, ncol = 8)
lgrid[1,] <- c("r", "l", "q", "s", "t", "z", "c", "a")
lgrid[2,] <- c("i", "v", "d", "z", "h", "l", "t", "p")
lgrid[3,] <- c("u", "r", "o", "y", "w", "c", "a", "c")
lgrid[4,] <- c("x", "r", "f", "n", "d", "p", "g", "v")
lgrid[5,] <- c("h", "j", "f", "f", "k", "h", "g", "m")
lgrid[6,] <- c("k", "y", "e", "x", "x", "g", "k", "i")
lgrid[7,] <- c("l", "q", "e", "q", "f", "u", "e", "b")
lgrid[8,] <- c("l", "s", "d", "h", "i", "k", "y", "n")

```
This function checks if the token coordinates are on a green square


```{r, echo = TRUE}
green_sqrs <- matrix(NA, nrow = 4, ncol = 2)
green_sqrs[1,] <- c(2, 6)
green_sqrs[2,] <- c(3, 7)
green_sqrs[3,] <- c(6, 2)
green_sqrs[4,] <- c(7, 3)

is.green <- function(g_coords){
  g <- FALSE
  if (g_coords[1] == green_sqrs[1, 1] && g_coords[2] == green_sqrs[1, 2] || 
      g_coords[1] == green_sqrs[2, 1] && g_coords[2] == green_sqrs[2, 2] ||
      g_coords[1] == green_sqrs[3, 1] && g_coords[2] == green_sqrs[3, 2] ||
      g_coords[1] == green_sqrs[4, 1] && g_coords[2] == green_sqrs[4, 2]) {
    g <- TRUE
  }
  return(g)
}
```


# **Part 1:**
Function to move the counter using the specified rules
Making sure if the token is on the edge it moves to a random square

```{r, echo = TRUE}
move_counter <- function(start_coords){
  x <- c(start_coords[1])
  y <- c(start_coords[2])
  if (x == 1 || x == 8 || y == 1 || y == 8) {
    finish_coords <- sample(x = 1:8, size = 2, replace = T) 
  }
  else {
    (adj_coords <- sample(x = -1:1, size = 2, replace = T)) 
    while (adj_coords == c(0, 0)) adj_coords <- sample(x = -1:1, size = 2, replace = T)
    finish_coords <- start_coords + adj_coords
  }
  
  return(finish_coords)
}
```

Forming function to decide whether to take a letter or not
```{r}
take_letter <- function(letters, next_letter) {
  take.letter <- sample(100, size = 1)
    if (take.letter > 50) {
      letters <- c(letters, next_letter)
    }
return(letters)
}


```


A function which checks if the letter collection forms a palindrome

```{r, echo = TRUE}
is.palindrome <- function(letters) {
    word <- paste(letters, sep = "", collapse = "")
    rawWord <- charToRaw(tolower(word)) ## converts to lower case
    palindrome <- identical(rawWord, rev(rawWord))
    return(palindrome)
}

```


# **Part 2:**

The player has a random 50% chance of adding the letter to the players collection,
until the fourth letter which is only picked up if it leads to less than four unique
letters in the players collection. Otherwise, it would be impossible to form a palindrome
when picking up a fifth letter. The fifth letter is only added if it causes the players
letter collection to form a palindrome.

An alternative would be to implement user input. For example,if the token lands on
a white square the user will have a choice of whether to collect that letter or not.
Once they reach the maximum number of letters (5) 
when landing on a white square they will be able to decide whether they want that letter.
If the player does they can then decide which letter from their collection to replace.



Forming function to check if adding a fifth letter forms a palindrome

```{r, echo = TRUE}
fifth_letter <- function(letters, next_letter){
  five_letters <- (c(letters, next_letter))
  palin <- sort_palindrome(five_letters)
  palindrome <- is.palindrome(palin)
  return(palindrome)
}

```

Function to ensure that a fourth letter is only picked up if it leads to less than
four unique letters in the players collection.

```{r}

fourth_letter <- function(letters, next_letter){
  four_letters <- (c(letters, next_letter))
  uniques <- unique(four_letters)
  if (length(uniques) < 4) {
    return(four_letters)
  }
  return(letters)
}
```


Function to sort the five letters into a palindrome if possible
This function is combined with the is.palindrome function to check if a palindrome
can be formed.
```{r}
sort_palindrome <- function(p_letters){
  uniques <- unique(p_letters)
  palin <- list()
  x <- 1
  for (u in uniques) {
    counts <- 0
    for (i in p_letters){
      if (i == u) counts <- counts + 1
    }
    while (counts > 0){
      if (counts %%2 == 1) {
        palin[3] <- u
        counts <- counts - 1
      }
      else {
        palin[x] <- u
        palin[6-x] <- u
        x <- x + 1
        counts <- counts - 2
          
      }  
    }
  }
return(palin)
}
```

# **Part 3:**
Function to simulate the game, returning the number of moves required.
```{r, echo = TRUE}

play_game <- function(lgrid, green_coords, start_coords, p){
  letters <- list()
  palindrome <- FALSE
  num_moves <- 1
  start_letter <- lgrid[start_coords[1], start_coords[2]]
  take_letter(letters, start_letter)
  
  while (palindrome == FALSE){
    next_coords <- c(move_counter(start_coords))
    start_coords <- next_coords
    num_moves <- num_moves + 1
    next_letter <- lgrid[next_coords[1], next_coords[2]]

    
    if (is.green(next_coords) == TRUE) {
      prob <- sample(100, size = 1) / 100 
      if (prob >= p) {
        letters <- list("f", "f", "h", "k")
      }
      else {
          letters <- letters[(letters) %in% next_letter == FALSE]
      }
    }
    
    else if (length(letters) == 3) {
      letters <- fourth_letter(letters, next_letter)
    }
    
    else if (length(letters) == 4) {
      palindrome <- fifth_letter(letters, next_letter)
      
      
      if (palindrome == TRUE) {
        letters <- (c(letters, next_letter))
        num_moves <- num_moves + 1
        # print(paste("Well done, you won in", num_moves, "moves!"))
        return(num_moves)
      }
  
    }
    else {
      letters <- take_letter(letters, next_letter)
    }
  }
}

```


# **Part 4:**

To find out how the time to finish depends on p a large number of realisations can
be simulated for different values of p. These realisations can be plotted & compared
to see how the mean & variance differ to show how the number of moves required to
form a palindrome is dependent on p.
10000 realisations will be assessed for 3 different values of p (0.9, 0.1, 0.5).


```{r, echo=FALSE, fig.height = 4, cache=TRUE}
set.seed(200206688)
start_coords <- c(4, 4)


p <- 0.1


low_p_num_moves<- replicate(10000, play_game(lgrid, green_coords, start_coords
                                            , p))
hist(low_p_num_moves, main ="Histogram of the number of moves to complete the
     game (p=0.1)",
     xlab = "Number of Moves" )



p <- 0.5

mid_p_num_moves<- replicate(10000, play_game(lgrid, green_coords, start_coords
                                            , p))
hist(mid_p_num_moves, main ="Histogram of the number of moves to complete the
     game (p=0.5)",
     xlab = "Number of Moves" )


p <- 0.9


high_p_num_moves <- replicate(10000, play_game(lgrid, green_coords, start_coords
                                              , p))
hist(high_p_num_moves, main ="Histogram of the number of moves to complete the 
     game (p=0.9) ",
     xlab = "Number of Moves" )



```
```{r, echo = FALSE}

low_p_mean <- mean(low_p_num_moves)
low_p_var <- var(low_p_num_moves)

mid_p_mean <- mean(mid_p_num_moves)
mid_p_var <- var(mid_p_num_moves)

high_p_mean <- mean(high_p_num_moves)
high_p_var <- var(high_p_num_moves)

low_p_var <- sprintf(low_p_var, fmt = '%#.3f')  
mid_p_var <- sprintf(mid_p_var, fmt = '%#.3f')  
high_p_var <- sprintf(high_p_var, fmt = '%#.3f')  

print(paste("Mean for 0.1p:", low_p_mean, " Mean for 0.5p:", mid_p_mean,
            "Mean for 0.9p:", high_p_mean))


print(paste("Variance for 0.1p:", low_p_var, " Variance for 0.5p:", mid_p_var,
            "Variance for 0.9p:", high_p_var))

```

These three values of p are compared over a large number of realisations and as shown
by their means there appears to be a degree of dependence on the value of p for the 
expected number of moves. With an increase in the value of p accompanied by an
increase in the average number of moves required. This is shown by the sample mean rising at 
each increment in the value of p.
Furthermore, the variance increases alongside an increasing value of p indicating a much
greater variation in the number of moves required for larger p values. Both these summary
statistics illustrate that there is a dependence on the value of p, with a larger value
leading to a higher number of moves required and a greater spread in the distribution.


# Part 5:

To investigate this hypothesis the game can be simulated 10000 times for each scenario
to determine whether the two scenarios can be assumed to be identical.

```{r, echo=FALSE, fig.height = 4, cache = TRUE}
set.seed(200206688)
start_coords <- c(4, 4)
p <- 0.95


d4_num_moves <- replicate(10000, play_game(lgrid, green_coords, start_coords, p))
hist(d4_num_moves, main ="Histogram of the number of moves to complete the game (D4)",
     xlab = "Number of Moves" )

start_coords <- c(6, 6)
p <- 0.05


f6_num_moves<- replicate(10000, play_game(lgrid, green_coords, start_coords, p))
hist(f6_num_moves, main ="Histogram of the number of moves to complete the game (F6)",
     xlab = "Number of Moves" )

```


```{r, echo=FALSE}

set.seed(200206688)
f6_mean <- mean(f6_num_moves)
f6_var <- var(f6_num_moves)
f6_low_q <- quantile(f6_num_moves, 0.025)
f6_high_q <- quantile(f6_num_moves, 0.975)
f6_CI <- c(f6_low_q, f6_high_q)

d4_mean <- mean(d4_num_moves)
d4_var <- var(d4_num_moves)
d4_low_q <- quantile(d4_num_moves, 0.025)
d4_high_q <- quantile(d4_num_moves, 0.975)
d4_CI <- c(d4_low_q, d4_high_q)
f6_var <- sprintf(f6_var, fmt = '%#.3f')  
d4_var <- sprintf(d4_var, fmt = '%#.3f')  

print(paste("Mean for D4 start position:", d4_mean, " Mean for F6 start position:", f6_mean))
print(paste("Variance for D4 start position:", d4_var, " Variance for F6 start position:", f6_var))

print(d4_CI)
print(f6_CI)
```

The evidence suggests that the probability distributions of the number of moves are not identical. This is due to a variety of factors, firstly, there is a notable deviation in the sample means for the two scenarios with the D4 start position having a larger average of 26.0 compared to 20.2 for the F6 start position. Furthermore, the D4 start position displays a much larger variance which relates to the larger p value. In part 4 it was shown that larger p values resulted in a larger variance for the number of moves required. Finally, the stark contrast in the 95% confidence intervals for the two scenarios reinforces the difference in probability distributions. The D4 data shows a confidence interval of 8-73 moves whilst the F6 data has a confidence interval of 8 to 48 moves. The difference between the two distributions is found in the upper interval with the D4 scenario more skewed towards higher values.


# **Part 6:**

The Monte Carlo approximation of the expectation of a random variable is the sample mean.
As a result, the expectation of the two squares can be assumed to be the sample mean of the
two samples.

```{r}
sqr_A <- c(25,13,16,24,11,12,24,26,15,19,34)
sqr_B <- c(35,41,23,26,18,15,33,42,18,47,21,26)
mean1 <- mean(sqr_A)
mean2 <- mean(sqr_B)
```

```{r, echo=FALSE}
mean1_2dp <- sprintf(mean1, fmt = '%#.2f')
print(paste("The mean number of moves for square A is:", mean1_2dp))
print(paste("The mean number of moves for square B is:", mean2))

```

As shown above the two samples display a large difference in expectation if using the 
Monte Carlo approximation. However, the null hypothesis that the expectation values are
equal for the two scenarios can be tested using randomization tests.

The p-value can be calculated using the function t.test as the hypothesis involves a
two_sided test.

```{r}
t.test(x = sqr_A, y = sqr_B, alternative = "two.sided", paired = F)


```
```{r}

T_obs <- -2.3463
p_value <- 0.02968

```



The test results can be randomly assigned to the two different groups with the test
statistic (p-value) recalculated. A similar p-value infers that the hypothesis is 
correct and the expectation values can be assumed to be equal.

```{r}

all_data <- c(25,13,16,24,11,12,24,26,15,19,34,35,41,23,26,18,15,33,42,18,47,21,26)

calc_T <- function(){
  perm <- sample(1:23,23, replace = F)
  sqr_A <- all_data[perm[1:11]]
  sqr_B <- all_data[perm[12:23]]
  mean1 <- mean(sqr_A)
  mean2 <- mean(sqr_B)
  s <- sqrt((var(sqr_A) * 7 +  var(sqr_B) * 7 ) / 14)
  (mean1 - mean2) / (s / 2)
}
n <- 10000
sim_T <- replicate(n - 1, calc_T())
p_val <- (sum(abs(sim_T) >= abs(T_obs)) + 1) / n


```

```{r, echo=FALSE}
print(paste("The original p value is", p_value))
print(paste("The new p value is:", p_val))

```


The two p-values are quite dissimilar, with the null hypothesis p-value much lower
than the original p-value. So as a result, using randomization tests the null 
hypothesis is rejected and the two expectation values are not assumed to equal.



